---
title: "WebApps 6502"
subtitle: "Session 1.2--R Shiny Basics"
footer: "[WebApps Website](https://rshiny-is.cool)"
logo: images/webapps-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    chalkboard: true
execute:
  echo: true
  message: false
  warning: false
  freeze: auto
---

# Overview

## Shiny App Basics

Two parts of a Shiny app:

- **UI**: User Interface
- **Server**: Server logic

They come together in a single script: `app.R`

::: aside
**Note:** You can compose your app with the ui and server in separate scripts, but we will stick to the single `app.R` script for now.
:::

## The app.R Script

```{r} 
#| label: app.r
#| code-line-numbers: "|1|3-5|7-8|10"
#| eval: false

library(shiny)

ui <- fluidPage(
  "Hello, Shiny!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

# The User Interface (`ui`)

## The UI

```{r} 
#| label: ui
#| code-line-numbers: "3-5"
#| eval: false

library(shiny)

ui <- fluidPage(
  "Hello, Shiny!"
)

server <- function(input, output, session) {
}

shinyApp(ui, server)
```

`fluidPage()` creates a Bootstrap page

## The UI

- [Bootstrap](https://getbootstrap.com/docs/5.0/getting-started/introduction/) is a free front-end framework used in web development  
- It is built on **HTML**, **CSS** and **JavaScript**
- There are many version of Bootstrap, the latest being version 5, but R Shiny uses version 3 by default

::: aside
With the `bslib` package you can use Bootstrap 4 or 5 in your R Shiny apps.
:::

## The UI

```{r}
#| echo: false

library(shiny)
```

Under the hood, every UI function is translated to HTML.

```{r}
as.character(titlePanel("Hello, Shiny!"))
```

<br>

This HTML can have styles attached to it, either by the `style` attribute, or by a CSS class.

```{r}
as.character(sidebarPanel())
```

## The UI

To let users interact with your app, you can add [input controls](https://shiny.posit.co/r/getstarted/shiny-basics/lesson3/). A basic input control has:

- an `inputId`
- a `label`
- a `value`

## The UI

For example:

```{.r}
numericInput(inputId = "number", 
             label = "Enter a number", 
             value = 0)
```
<br>
```{.r}
actionButton(inputId = "button", 
             label = "Click me")
```

## Updating the UI

::: {.incremental}
- When the app runs, every user gets served the same HTML from the UI
- When the user interacts with the UI, we want that HTML to **react** and update based on user input
- That is possible because these outputs are **reactive outputs**
- The server logic (which we will discuss in detail soon) uses **reactive programming** that updates related inputs in the UI
:::

## Updating the UI

Shiny has several functions that turn R objects into reactive outputs for your `ui`: the [Output](https://shiny.posit.co/r/getstarted/shiny-basics/lesson4/) family of functions.

Each function creates a specific type of output, for example:

| UI Function | Output type |
|----------|-------------|
| textOutput() | text |
| tableOutput() | table |
| plotOutput() | plot |
| uiOutput() | raw HTML |
: {.table .table-striped .table-hover}

## Updating the UI

Every output element needs a single argument: `outputId`. This is a simple string that needs to be **unique**.

<br>

```{.r}
textOutput(outputId = "text")
```

<br>

This `textOutput` tells Shiny what to display. It's a **placeholder** for what is goign to be produced by the server logic.

# The Server Logic

## The Server

::: {.incremental}
- The server function builds an object named **output** and this object will contain all the code needed to display the R objects in the UI
- This output object is **list-like**
- Each R object that you want to display has its own entry in this list
- Therefore the name of every output in your UI needs to match a definition in the server, e.g. `output$text`
:::

## The Server

Each entry in the output list should contain a **Render** function. You must use **Render** functions that match the **Output** functions:

| UI Function | Output type | Render function |
|----------|-------------|-----------------|
| textOutput() | text | renderText() |
| tableOutput() | table | renderTable() |
| plotOutput() | plot | renderPlot() |
| uiOutput() | raw HTML | renderUI() |
: {.table .table-striped .table-hover}

## The Server

```{r} 
#| code-line-numbers: "2,6"

ui <- fluidPage(
  textOutput(outputId = "text")
)

server <- function(input, output, session) {
  output$text <- renderText({
  })
}
```

## Using Inputs

- This works the same with inputs.
- The server can access an object named `input` that will contain the value of all the input objects defined in the UI.
- This object is **list-like**, just like the output object. Each input value has its own entry in this list, e.g. `input$number`.

## 

Combining the input and output objects, we get a simple app that displays the square of a number. ðŸ‘

```{r}
#| eval: false
library(shiny)

ui <- fluidPage(
  numericInput(inputId = "number", 
               label = "Enter a number", 
               value = 0),
               
  textOutput(outputId = "text")
)

server <- function(input, output, session) {
  output$text <- renderText({
    input$number^2
  })
}

shinyApp(ui, server)
```

## Your Turn! 

- Create a Shiny app that displays the square of a number
- Try entering different numbers manually
- Try using the up and down arrows
- Adjust the server logic to display the cube of a number
- Try other calculations

```{r}
#| echo: false

library(countdown)
countdown(minutes = 10, 
          id = "timer1", 
          bottom = "10%",
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

## Eager Output

An output is **eager**: it will update as soon as the input changes.

<br>

This **eagerness** is handy: you don't need to worry about updating the output when the input changes.

<br>

But what if you want to trigger the calculation only when you want?

## Adding a Reactivity Modifier

You could use an `actionButton` as an **event**:

```{r} 
#| code-line-numbers: "8-9,15-17"

library(shiny)

ui <- fluidPage(
  numericInput(inputId = "number", 
               label = "Enter a number", 
               value = 0),
               
  actionButton(inputId = "button", 
               label = "Click me"),
               
  textOutput(outputId = "text")
)

server <- function(input, output, session) {
  output$text <- renderText({
    input$number^2
  }) |> bindEvent(input$button)
}

shinyApp(ui, server)
```

## Try It Yourself! 

- Add an `actionButton` to your app
- Adjust the text on the button to read "Calculate"
- Make sure it functions as intended
- Go to the [getting started](https://shiny.posit.co/r/getstarted/shiny-basics/lesson3/) lesson on control widgets
- See if you can control the calculation with a checkbox instead of a button
- What is wrong with this approach?

## Acknowledgements

- Parts of this presentation were adapted from Veerle van Leemput's [Shiny 101: The Modular App Blueprint](https://github.com/hypebright/shinyconf2024-shiny101)
